### 2025-08-22 chapters/background.tex
citations & plot amendments are complete for chapters 1 - 2

The Keystone framework is founded on the RISC-V instruction set architecture (ISA), which provides a compelling platform for secure system design due to its open specification and extensible modularity~\cite{lee2019keystone,dayeol2019keystone}. Unlike traditional ISAs governed by proprietary vendors, RISC-V fosters a collaborative ecosystem that supports community-driven enhancements and custom extensions. This openness uniquely positions RISC-V as a prime candidate for trusted computing research and development~\cite{Survey2023}. Furthermore, its well-defined privilege model facilitates the construction of robust isolation mechanisms, which are essential for implementing Trusted Execution Environments (TEEs).\footnote{For an in-depth exposition of RISC-V privilege modes and PMP, see the RISC-V Privilege Specification v1.10~\cite{keystone2025how}.}

RISC-V specifies three hierarchical privilege levels: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode). M-mode operates with the highest privilege and possesses unrestricted access to hardware resources, managing critical functions such as interrupt handling, exception processing, and physical memory protection~\cite{lee2019keystone}. Keystone capitalizes on this privilege hierarchy by situating its Security Monitor (SM)—the core Trusted Computing Base (TCB)—within M-mode. The SM is responsible for managing enclaves, enforcing strict memory isolation, and mediating access to sensitive operations. By anchoring the TCB at this highest privilege level, Keystone minimizes the trusted surface exposed to potential attackers.

A key hardware feature leveraged by Keystone is the Physical Memory Protection (PMP) unit, introduced in the RISC-V Privilege Specification v1.10. PMP empowers M-mode software to define fine-grained access permissions for physical memory regions, specifying which lower privilege levels (U and S) may read, write, or execute code therein~\cite{keystone2025how}. This hardware-enforced mechanism forms the cornerstone of Keystone’s memory isolation strategy. Each enclave is assigned a dedicated, PMP-protected memory region that remains inaccessible to the host OS or unauthorized software, guaranteeing strong confidentiality and integrity guarantees enforced at the hardware level~\cite{lee2019keystone}.

In addition to PMP, RISC-V’s flexible trap and exception handling model enables M-mode to intercept all traps while optionally delegating certain classes, such as system calls and page faults, to S-mode for efficiency~\cite{lee2019keystone}. This delegation mechanism facilitates cooperation between enclave runtimes and the host OS for memory management without compromising enclave boundaries. Keystone utilizes the RISC-V Memory Management Unit (MMU) to support private virtual address spaces per enclave, further fortified by PMP protections against unauthorized access or tampering~\cite{dayeol2019keystone}.

Notably, the Keystone Security Monitor is implemented in C using standard programming toolchains, which enhances maintainability, auditability, and formal verification potential. This contrasts with fixed-function microcode or proprietary firmware that is often opaque and difficult to analyze, thus bolstering Keystone’s transparency and security assurance~\cite{lee2019keystone}.

@@

A TEE effectively partitions system resources such as CPU registers, memory, and I/O peripherals, restricting access exclusively to trusted enclave applications. This isolation is enforced by the processor’s privilege levels, memory protection units, or dedicated security monitors, thereby preventing unauthorized entities—including kernel-level malware or user-space adversaries—from accessing or manipulating enclave data~\cite{brasser2019software}.

Security in TEEs hinges on minimizing the Trusted Computing Base (TCB), which encompasses all components that must be trusted for correct and secure operation. A smaller TCB reduces the risk of vulnerabilities and facilitates formal verification and security audits~\cite{lee2019keystone,Survey2023}. Modern TEEs strive to encapsulate only essential components within the TCB, balancing security with functionality.

Commercially dominant TEE solutions include Intel Software Guard Extensions (SGX) and ARM TrustZone. Intel SGX enables enclave creation at user privilege levels, employing hardware-enforced memory encryption and access controls~\cite{costan2016intel}. ARM TrustZone employs a split-world model that partitions processor execution into secure and normal worlds, with transitions mediated by secure monitor calls~\cite{yan2018trustzone}. Despite their widespread deployment, these architectures suffer from proprietary designs and limited transparency, which constrain their flexibility for academic research and system-level innovation.

@@

In contrast, open-source TEE frameworks such as Keystone have emerged to address these limitations by leveraging the open, modular RISC-V ISA~\cite{lee2019keystone,dayeol2019keystone}. The modularity and openness of RISC-V facilitate the design of TEEs that are both highly configurable and more transparent, fostering security research and experimentation~\cite{Survey2023}. A central architectural principle of TEEs, exemplified by Keystone, is privilege separation: executing the TCB in the highest privilege mode (M-mode in RISC-V) while relegating untrusted software to lower privilege levels. This design enforces strict isolation policies, preventing untrusted software from compromising enclave security~\cite{lee2019keystone}.

\begin{table}[htbp]
\centering
\small
%\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}p{3cm}p{2.5cm}p{2.5cm}p{2.5cm}@{}}
\toprule
\textbf{Feature} & \textbf{Intel SGX} & \textbf{ARM TrustZone} & \textbf{Keystone} \\
\midrule
Privilege Model & Ring 3 enclaves & Secure Monitor & M-mode TCB \\
Openness & Closed & Closed & Open \\
Use Cases & Cloud security & Mobile security & Research, IoT \\
Hardware Dependency & Intel CPUs & ARM SoCs & RISC-V \\
\bottomrule
\end{tabular}
\caption{Comparison of three TEE implementations.}
\end{table}

@@

Keystone is an open-source Trusted Execution Environment (TEE) framework built on the RISC-V architecture, designed to enable customizable, lightweight, and secure enclaves suitable for both research and deployment in resource-constrained or security-critical settings \cite{dayeol2019keystone}. The framework’s modular design emphasizes a minimal Trusted Computing Base (TCB), portability across diverse RISC-V platforms, and extensibility to accommodate advanced security features, reflecting a clean and flexible architectural philosophy.

Keystone leverages the multiple privilege modes defined by the RISC-V specification—user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode)—to enforce security policies effectively. At the core of Keystone’s TCB is the Security Monitor (SM), which executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and access control enforcement. The SM uses RISC-V’s Physical Memory Protection (PMP) mechanism to isolate enclave memory regions dynamically, ensuring strict access controls during enclave execution \cite{dayeol2019keystone}.

A typical Keystone-enabled platform comprises RISC-V cores augmented with secure boot mechanisms, trusted entropy sources, and optionally, hardware accelerators for cryptographic functions or secure I/O. Importantly, Keystone does not require hardware beyond standard RISC-V compliance and PMP support, making it adaptable to implementations ranging from FPGA prototypes to silicon SoCs \cite{dayeol2019keystone}.

Each enclave in Keystone consists of two main software components: a user-level enclave application (eapp) containing application-specific logic, and a supervisor-level runtime (RT) providing essential operating system abstractions such as exception handling, system call services, and memory management. This layered approach reduces the complexity of enclave applications and provides a trusted runtime environment isolated from the untrusted operating system and other processes \cite{dayeol2019keystone}.

The enclave lifecycle is structured into three primary phases: creation, execution, and destruction. During creation, the SM validates and locks the enclave’s memory layout—referred to as Enclave Page Memory (EPM)—with PMP to ensure confidentiality and integrity. During execution, the SM mediates context switches between the host and enclave, dynamically adjusting PMP to maintain isolation. Destruction securely erases enclave memory to prevent any residual data leakage \cite{dayeol2019keystone}. Keystone also supports optional features such as remote attestation, allowing external verifiers to authenticate enclave state prior to provisioning sensitive data, and can be extended for secure I/O and cryptographic acceleration.

@@

Keystone’s modular and open framework leverages the openness of RISC-V to provide strong security guarantees while enabling flexibility for research and customization. Its minimal TCB, modular layering (Security Monitor, runtime, eapp), and standards-compliant interfaces make it an attractive platform for developing and evaluating secure applications on modern processors \cite{dayeol2019keystone}. 
To support secure communication and attestation, Keystone enclaves can integrate modern cryptographic primitives, including emerging post-quantum schemes such as Kyber.

@@

In RISC-V, the PMP mechanism is configured exclusively by machine-mode (M-mode) software—the most privileged execution level in the architecture—and applies access control policies to all lower-privileged software running in supervisor (S-mode) or user mode (U-mode). Each PMP entry defines a memory region along with a set of permissions (read, write, execute) and an addressing mode. The hardware enforces these permissions such that any access attempt by S-mode or U-mode software that violates a PMP rule results in a trap to M-mode. This mechanism ensures that M-mode retains full control over physical memory while being able to delegate controlled access to untrusted software layers~\cite{riscvprivspec,lee2019keystone}.

@@

In the context of TEEs such as the Keystone framework, PMP serves as the fundamental enforcement mechanism for enclave isolation. Keystone leverages PMP to assign a protected memory region, known as the Enclave Page Memory (EPM), to each enclave. This region is configured to be inaccessible to both the untrusted operating system (running in S-mode) and any other software outside the enclave context~\cite{dayeol2019keystone}. During enclave creation, the Keystone Security Monitor (SM)—executing in M-mode—programmatically sets PMP entries to lock the memory region associated with the enclave. The SM verifies the integrity of the enclave binary, computes its cryptographic hash, and sets up the memory layout in such a way that only the enclave code has access to its memory contents~\cite{lee2019keystone,keystone2025how}.

@@

Furthermore, PMP entries are reconfigured dynamically during enclave context switches. For example, when the host operating system regains control, the PMP settings are altered to revoke access to the enclave’s memory. Conversely, when the enclave resumes execution, the SM reinstates its PMP region. This dynamic switching is essential to enforce temporal isolation and prevent time-of-check-to-time-of-use (TOCTTOU) vulnerabilities that may arise during state transitions~\cite{cerberus2022}.

Despite its utility, PMP has several practical limitations. The fixed number of PMP entries restricts the number of simultaneously protected regions, posing a challenge for complex systems with many isolated domains or subregions. Moreover, overlapping PMP regions are resolved based on entry priority (lower-numbered entries take precedence), requiring careful configuration to avoid inadvertent leaks or privilege escalations~\cite{riscvprivspec}.

@@

Security researchers have explored various enhancements and formal models to increase the reliability and verifiability of PMP-based isolation. For example, Cerberus~\cite{cerberus2022} builds a formally verified memory-sharing abstraction over PMP to enable safe communication between enclaves and untrusted components. Similarly, HexBox~\cite{hexbox2018} leverages PMP for real-time embedded security by associating memory regions with specific control-flow paths in embedded firmware, demonstrating PMP's applicability beyond general-purpose computing.

\begin{table}[htbp]
\centering
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{>{\raggedright}p{3cm} >{\raggedright}p{4cm} >{\raggedright\arraybackslash}p{5cm}}
\hline
\textbf{Limitation} & \textbf{Impact} & \textbf{Enhancement / Research} \\
\hline
Fixed \# entries & Limits scalability & Cerberus verified sharing \\
No dynamic resizing & Hard to adapt to workloads & Dynamic PMP proposals \\
Limited granularity & Coarse protection & Fine‑grained tagging \\
\hline
\end{tabular}
\caption{Summary of PMP limitations and research‑driven enhancements.}
\label{tab:pmp-limitations}
\end{table}

The role of PMP in Keystone exemplifies how low-complexity hardware primitives can be orchestrated by secure software layers to implement strong isolation guarantees. This synergy is particularly effective in the RISC-V ecosystem, where the open architecture allows deep customization, formal verification, and academic experimentation~\cite{Survey2023}. The combination of PMP, a secure boot mechanism, and a minimal TCB enables Keystone to enforce hardware-rooted enclave isolation without relying on opaque, vendor-specific extensions, setting it apart from alternatives like Intel SGX or ARM TrustZone~\cite{dayeol2019keystone,turn0search0}.

In conclusion, Physical Memory Protection in RISC-V is a powerful, lightweight, and deterministic mechanism that underpins secure execution in many RISC-V-based platforms. Its integration into frameworks like Keystone enables transparent and verifiable enclave isolation while remaining suitable for constrained systems. As the RISC-V ecosystem continues to evolve, PMP will remain a fundamental building block for secure, open, and formally analyzable trusted computing platforms.

@@

Kyber is a post-quantum key encapsulation mechanism (KEM) designed to resist attacks from both classical and quantum adversaries. Its security is based on the hardness of the Module Learning With Errors (Module-LWE) problem, a structured lattice problem considered intractable even for quantum computers. Kyber is one of the finalists in the NIST Post-Quantum Cryptography (PQC) competition and has been selected for standardization due to its strong security and performance characteristics \cite{kyber2024}.  % [SOURCE1]

At its core, Kyber is derived from the Lyubashevsky–Peikert–Regev (LPR) encryption framework, which adapts Learning With Errors (LWE) from integer vectors to more efficient polynomial structures. Specifically, Kyber operates in the Module-LWE setting, which balances the structure of Ring-LWE with the flexibility of traditional LWE. This results in strong performance and compact key sizes while preserving security \cite{kyber2024,kyber2021}. % [SOURCE1, BLOG1]

Kyber supports three security levels:
\begin{itemize}
    \item \textbf{Kyber-512}, targeting AES-128-level security,
    \item \textbf{Kyber-768}, targeting AES-192-level security (recommended),
    \item \textbf{Kyber-1024}, targeting AES-256-level security.
\end{itemize}

\begin{figure}[htbp]
\centering
\includegraphics[width=0.9\linewidth]{figures/kyber_768.png}
\caption{Kyber768 Key Exchange Algorithm}
\label{fig:kyber_768}
\end{figure}

\begin{table}[h]
\centering
\caption{Kyber and RSA: Key and Ciphertext Sizes (in bytes)}
\label{tab:kyber_sizes}
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Scheme} & \textbf{Private Key} & \textbf{Public Key} & \textbf{Ciphertext} \\
\hline
Kyber-512 & 1632 & 800 & 768 \\
Kyber-768 & 2400 & 1184 & 1088 \\
Kyber-1024 & 3168 & 1568 & 1568 \\
RSA-3072 & 384 & 384 & 384 \\
RSA-15360 & 1920 & 1920 & 1920 \\
\hline
\end{tabular}
\end{table}

While RSA keys are somewhat smaller, Kyber maintains significantly better security in the post-quantum context, and its keys are far more compact than those of many alternative PQC candidates such as Classic McEliece \cite{kyber2021}. % [BLOG1]

Kyber is structured as a KEM rather than a traditional public-key encryption (PKE) scheme. Internally, however, it uses a PKE system transformed into a CCA-secure KEM. The KEM exposes a simple API:
\begin{itemize}
    \item $(\mathit{pk}, \mathit{sk}) = \texttt{KeyGen}()$
    \item $(\mathit{ct}, \mathit{ss}) = \texttt{Encapsulate}(\mathit{pk})$
    \item $\mathit{ss} = \texttt{Decapsulate}(\mathit{sk}, \mathit{ct})$
\end{itemize}

The encapsulation mechanism makes Kyber especially suitable for protocols like TLS, Signal, and OpenPGP. % [BLOG1]

A simplified pedagogical version called ``Baby Kyber’’ is often used to illustrate the cryptographic structure. Baby Kyber uses polynomials over a small modulus and limited degree to show how private and public keys, as well as ciphertexts, are constructed and manipulated. Encryption relies on adding small random errors to hide the message, while decryption relies on error tolerance due to coefficient scaling. If the message polynomial is scaled appropriately, decryption can recover the original bit string by rounding \cite{kyber2021}. % [BLOG1]

In full Kyber, the parameters are chosen to ensure security and correctness:
\begin{itemize}
    \item $n = 256$: polynomial degree
    \item $k \in \{2, 3, 4\}$: number of polynomials (Kyber-512, 768, 1024)
    \item $q = 3329$: prime modulus
    \item $\eta_1, \eta_2$: bounds on error distributions
    \item $d_u, d_v$: compression factors
\end{itemize}

Kyber’s performance has been optimized for a variety of platforms. Table~\ref{tab:kyber_perf} shows benchmark results on an Intel Haswell CPU with AVX2 vectorization \cite{kyber2024}. % [SOURCE1]

\begin{table}[h]
\centering
\caption{Kyber Performance on Haswell CPU (AVX2 Optimized)}
\label{tab:kyber_perf}
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Variant} & \textbf{KeyGen (cycles)} & \textbf{Encaps (cycles)} & \textbf{Decaps (cycles)} \\
\hline
Kyber-512 & 33,856 & 45,200 & 34,572 \\
Kyber-768 & 52,732 & 67,624 & 53,156 \\
Kyber-1024 & 73,544 & 97,324 & 79,128 \\
\hline
\end{tabular}
\end{table}

Kyber also includes a variant called \texttt{Kyber-90s}, which replaces the use of SHAKE with AES-256 and SHA2 primitives, enabling further acceleration on platforms with AES instruction support \cite{kyber2024}. % [SOURCE1]

The security of Kyber ultimately reduces to the hardness of the Shortest Vector Problem (SVP) via the Module-LWE problem. Recovering a message or key requires solving an instance of SVP in a structured lattice, a task that remains intractable even with quantum resources\cite{kyber2021}. % [BLOG1]

>>>>

The Keystone framework is founded on the RISC-V instruction set architecture (ISA), which provides a compelling platform for secure system design due to its open specification and extensible modularity~\cite{Lee2019,Lee2019}. Unlike traditional ISAs governed by proprietary vendors, RISC-V fosters a collaborative ecosystem that supports community-driven enhancements and custom extensions. This openness uniquely positions RISC-V as a prime candidate for trusted computing research and development~\cite{Survey2023}. Furthermore, its well-defined privilege model facilitates the construction of robust isolation mechanisms, which are essential for implementing Trusted Execution Environments (TEEs).\footnote{For an in-depth exposition of RISC-V privilege modes and PMP, see the RISC-V Privilege Specification v1.10~\cite{keystone2025how}.}

RISC-V specifies three hierarchical privilege levels: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode). M-mode operates with the highest privilege and possesses unrestricted access to hardware resources, managing critical functions such as interrupt handling, exception processing, and physical memory protection~\cite{Lee2019}. Keystone capitalizes on this privilege hierarchy by situating its Security Monitor (SM)—the core Trusted Computing Base (TCB)—within M-mode. The SM is responsible for managing enclaves, enforcing strict memory isolation, and mediating access to sensitive operations. By anchoring the TCB at this highest privilege level, Keystone minimizes the trusted surface exposed to potential attackers.

A key hardware feature leveraged by Keystone is the Physical Memory Protection (PMP) unit, introduced in the RISC-V Privilege Specification v1.10. PMP empowers M-mode software to define fine-grained access permissions for physical memory regions, specifying which lower privilege levels (U and S) may read, write, or execute code therein~\cite{keystone2025how}. This hardware-enforced mechanism forms the cornerstone of Keystone’s memory isolation strategy. Each enclave is assigned a dedicated, PMP-protected memory region that remains inaccessible to the host OS or unauthorized software, guaranteeing strong confidentiality and integrity guarantees enforced at the hardware level~\cite{Lee2019}.

In addition to PMP, RISC-V’s flexible trap and exception handling model enables M-mode to intercept all traps while optionally delegating certain classes, such as system calls and page faults, to S-mode for efficiency~\cite{Lee2019}. This delegation mechanism facilitates cooperation between enclave runtimes and the host OS for memory management without compromising enclave boundaries. Keystone utilizes the RISC-V Memory Management Unit (MMU) to support private virtual address spaces per enclave, further fortified by PMP protections against unauthorized access or tampering~\cite{Lee2019}.

Notably, the Keystone Security Monitor is implemented in C using standard programming toolchains, which enhances maintainability, auditability, and formal verification potential. This contrasts with fixed-function microcode or proprietary firmware that is often opaque and difficult to analyze, thus bolstering Keystone’s transparency and security assurance~\cite{Lee2019}.

@@

A TEE effectively partitions system resources such as CPU registers, memory, and I/O peripherals, restricting access exclusively to trusted enclave applications. This isolation is enforced by the processor’s privilege levels, memory protection units, or dedicated security monitors, thereby preventing unauthorized entities—including kernel-level malware or user-space adversaries—from accessing or manipulating enclave data~\cite{brasser2020enclave}.

Security in TEEs hinges on minimizing the Trusted Computing Base (TCB), which encompasses all components that must be trusted for correct and secure operation. A smaller TCB reduces the risk of vulnerabilities and facilitates formal verification and security audits~\cite{Lee2019,Survey2023}. Modern TEEs strive to encapsulate only essential components within the TCB, balancing security with functionality.

Commercially dominant TEE solutions include Intel Software Guard Extensions (SGX) and ARM TrustZone. Intel SGX enables enclave creation at user privilege levels, employing hardware-enforced memory encryption and access controls~\cite{costan2016intel}. ARM TrustZone employs a split-world model that partitions processor execution into secure and normal worlds, with transitions mediated by secure monitor calls~\cite{yan2018trustzone}. These architectural designs are illustrated in Figure~\ref{fig:tee-architectures}, which compares TrustZone's dual-world architecture (Figure~\ref{fig:trustzone}) with SGX's Enclave Page Cache memory model (Figure~\ref{fig:sgx-epc}). Despite their widespread deployment, these architectures suffer from proprietary designs and limited transparency, which constrain their flexibility for academic research and system-level innovation.

@@

In contrast, open-source TEE frameworks such as Keystone have emerged to address these limitations by leveraging the open, modular RISC-V ISA~\cite{Lee2019}. The modularity and openness of RISC-V facilitate the design of TEEs that are both highly configurable and more transparent, fostering security research and experimentation~\cite{Survey2023}. A central architectural principle of TEEs, exemplified by Keystone, is privilege separation: executing the TCB in the highest privilege mode (M-mode in RISC-V) while relegating untrusted software to lower privilege levels. This design enforces strict isolation policies, preventing untrusted software from compromising enclave security~\cite{Lee2019}.
Table~\ref{tab:tee-comparison} summarizes key architectural and design distinctions between Intel SGX, ARM TrustZone, and Keystone.

\begin{table}[htbp]
\centering
\caption{Comparison of three TEE implementations.}
\label{tab:tee-comparison}
\small
%\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}p{3cm}p{2.5cm}p{2.5cm}p{2.5cm}@{}}
\toprule
\textbf{Feature} & \textbf{Intel SGX} & \textbf{ARM TrustZone} & \textbf{Keystone} \\
\midrule
Privilege Model & Ring 3 enclaves & Secure Monitor & M-mode TCB \\
Openness & Closed & Closed & Open \\
Use Cases & Cloud security & Mobile security & Research, IoT \\
Hardware Dependency & Intel CPUs & ARM SoCs & RISC-V \\
\bottomrule
\end{tabular}
\end{table}

@@

Keystone is an open-source Trusted Execution Environment (TEE) framework built on the RISC-V architecture, designed to enable customizable, lightweight, and secure enclaves suitable for both research and deployment in resource-constrained or security-critical settings \cite{Lee2019}. The framework’s modular design emphasizes a minimal Trusted Computing Base (TCB), portability across diverse RISC-V platforms, and extensibility to accommodate advanced security features, reflecting a clean and flexible architectural philosophy.

Keystone leverages the multiple privilege modes defined by the RISC-V specification—user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode)—to enforce security policies effectively. At the core of Keystone’s TCB is the Security Monitor (SM), which executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and access control enforcement. The SM uses RISC-V’s Physical Memory Protection (PMP) mechanism to isolate enclave memory regions dynamically, ensuring strict access controls during enclave execution \cite{Lee2019}.

A typical Keystone-enabled platform comprises RISC-V cores augmented with secure boot mechanisms, trusted entropy sources, and optionally, hardware accelerators for cryptographic functions or secure I/O. Importantly, Keystone does not require hardware beyond standard RISC-V compliance and PMP support, making it adaptable to implementations ranging from FPGA prototypes to silicon SoCs \cite{Lee2019}.

Each enclave in Keystone consists of two main software components: a user-level enclave application (eapp) containing application-specific logic, and a supervisor-level runtime (RT) providing essential operating system abstractions such as exception handling, system call services, and memory management. This layered approach reduces the complexity of enclave applications and provides a trusted runtime environment isolated from the untrusted operating system and other processes \cite{Lee2019}.

The enclave lifecycle is structured into three primary phases: creation, execution, and destruction. During creation, the SM validates and locks the enclave’s memory layout—referred to as Enclave Page Memory (EPM)—with PMP to ensure confidentiality and integrity. During execution, the SM mediates context switches between the host and enclave, dynamically adjusting PMP to maintain isolation. Destruction securely erases enclave memory to prevent any residual data leakage \cite{Lee2019}. Keystone also supports optional features such as remote attestation, allowing external verifiers to authenticate enclave state prior to provisioning sensitive data, and can be extended for secure I/O and cryptographic acceleration.
Figure~\ref{fig:keystone_grouped} illustrates the Keystone enclave design. Specifically, Figure~\ref{fig:keystone_overview} shows the architectural components, including the Security Monitor, enclave runtime, and the RISC-V privilege hierarchy. Figure~\ref{fig:enclave_lifecycle} depicts the enclave lifecycle phases: creation, execution, and destruction.

@@

Keystone’s modular and open framework leverages the openness of RISC-V to provide strong security guarantees while enabling flexibility for research and customization. Its minimal TCB, modular layering (Security Monitor, runtime, eapp), and standards-compliant interfaces make it an attractive platform for developing and evaluating secure applications on modern processors \cite{Lee2019}. 
To support secure communication and attestation, Keystone enclaves can integrate modern cryptographic primitives, including emerging post-quantum schemes such as Kyber.

@@

In RISC-V, the PMP mechanism is configured exclusively by machine-mode (M-mode) software—the most privileged execution level in the architecture—and applies access control policies to all lower-privileged software running in supervisor (S-mode) or user mode (U-mode). Each PMP entry defines a memory region along with a set of permissions (read, write, execute) and an addressing mode. The hardware enforces these permissions such that any access attempt by S-mode or U-mode software that violates a PMP rule results in a trap to M-mode. This mechanism ensures that M-mode retains full control over physical memory while being able to delegate controlled access to untrusted software layers~\cite{riscvprivspec,Lee2019}.
The basic hardware architecture of PMP within the RISC-V privilege hierarchy is illustrated in Figure~\ref{fig:pmp-arch}.

@@

In the context of TEEs such as the Keystone framework, PMP serves as the fundamental enforcement mechanism for enclave isolation. Keystone leverages PMP to assign a protected memory region, known as the Enclave Page Memory (EPM), to each enclave. This region is configured to be inaccessible to both the untrusted operating system (running in S-mode) and any other software outside the enclave context~\cite{Lee2019}. During enclave creation, the Keystone Security Monitor (SM)—executing in M-mode—programmatically sets PMP entries to lock the memory region associated with the enclave. The SM verifies the integrity of the enclave binary, computes its cryptographic hash, and sets up the memory layout in such a way that only the enclave code has access to its memory contents~\cite{Lee2019,keystone2025how}.
As shown in Figure~\ref{fig:pmp-keystone}, the PMP unit mediates memory access between the OS and the enclave to enforce isolation guarantees.

@@

Furthermore, PMP entries are reconfigured dynamically during enclave context switches. For example, when the host operating system regains control, the PMP settings are altered to revoke access to the enclave’s memory. Conversely, when the enclave resumes execution, the SM reinstates its PMP region. This dynamic switching is essential to enforce temporal isolation and prevent time-of-check-to-time-of-use (TOCTTOU) vulnerabilities that may arise during state transitions~\cite{lee2022cerberus}.

Despite its utility, PMP has several practical limitations. The fixed number of PMP entries restricts the number of simultaneously protected regions, posing a challenge for complex systems with many isolated domains or subregions. Moreover, overlapping PMP regions are resolved based on entry priority (lower-numbered entries take precedence), requiring careful configuration to avoid inadvertent leaks or privilege escalations (see Figure~\ref{fig:pmp-priority})~\cite{riscvprivspec}.

@@

Security researchers have explored various enhancements and formal models to increase the reliability and verifiability of PMP-based isolation. For example, Cerberus~\cite{lee2022cerberus} builds a formally verified memory-sharing abstraction over PMP to enable safe communication between enclaves and untrusted components. Similarly, HexBox~\cite{hexbox2018} leverages PMP for real-time embedded security by associating memory regions with specific control-flow paths in embedded firmware, demonstrating PMP's applicability beyond general-purpose computing.
Table~\ref{tab:pmp-limitations} summarizes key limitations in PMP and highlights recent research directions aimed at overcoming them.

\begin{table}[htbp]
\centering
\caption{Summary of PMP limitations and research‑driven enhancements.}
\label{tab:pmp-limitations}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{>{\raggedright}p{3cm} >{\raggedright}p{4cm} >{\raggedright\arraybackslash}p{5cm}}
\hline
\textbf{Limitation} & \textbf{Impact} & \textbf{Enhancement / Research} \\
\hline
Fixed \# entries & Limits scalability & Cerberus verified sharing \\
No dynamic resizing & Hard to adapt to workloads & Dynamic PMP proposals \\
Limited granularity & Coarse protection & Fine‑grained tagging \\
\hline
\end{tabular}
\end{table}

The role of PMP in Keystone exemplifies how low-complexity hardware primitives can be orchestrated by secure software layers to implement strong isolation guarantees. This synergy is particularly effective in the RISC-V ecosystem, where the open architecture allows deep customization, formal verification, and academic experimentation~\cite{Survey2023}. The combination of PMP, a secure boot mechanism, and a minimal TCB enables Keystone to enforce hardware-rooted enclave isolation without relying on opaque, vendor-specific extensions, setting it apart from alternatives like Intel SGX or ARM TrustZone~\cite{Lee2019,suzaki2021tsperf}.

Physical Memory Protection in RISC-V is a powerful, lightweight, and deterministic mechanism that underpins secure execution in many RISC-V-based platforms. Its integration into frameworks like Keystone enables transparent and verifiable enclave isolation while remaining suitable for constrained systems. As the RISC-V ecosystem continues to evolve, PMP will remain a fundamental building block for secure, open, and formally analyzable trusted computing platforms.

@@

Kyber is a post-quantum key encapsulation mechanism (KEM) designed to remain secure against both classical and quantum adversaries. It is based on the hardness of the Module Learning With Errors (Module-LWE) problem, a structured lattice problem believed to be difficult to solve even with quantum algorithms \cite{kyber2021}. The algorithm has been selected by NIST for standardization in its post-quantum cryptography project, reflecting both its strong theoretical foundation and practical efficiency \cite{kyber2024}.

Kyber uses polynomial arithmetic in a modular ring to perform its cryptographic operations. It is based on the LPR encryption framework \cite{kyber2021}, which adapts the Learning With Errors (LWE) problem from integer vectors to more structured polynomial rings. This allows for faster computation and more compact key sizes compared to classical public-key systems.

Kyber supports three variants, each corresponding to a different level of security roughly equivalent to asymmetric encryption:

\begin{verbatim}
Kyber-512   → AES-128-level security
Kyber-768   → AES-192-level security (recommended)
Kyber-1024  → AES-256-level security
\end{verbatim}

These parameters determine the number of polynomials, noise levels, and compression factors used internally. Figure~\ref{fig:kyber_768} illustrates the Kyber-768 key exchange process, in which two parties derive a shared secret using public key encapsulation and decapsulation.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.9\linewidth]{figures/kyber_768.png}
\caption{Kyber-768 Key Exchange Algorithm.}
\label{fig:kyber_768}
\end{figure}

Kyber exposes a compact and simple interface for use in protocols. Its API supports three core operations: key generation, encapsulation, and decapsulation, as shown in Listing~\ref{lst:kem_api}.

\begin{lstlisting}[style=kemstyle, caption={Kyber KEM API}, label={lst:kem_api}]
(pk, sk)    = KeyGen()           # Generate key pair
(ct, ss)    = Encapsulate(pk)    # Encrypt shared secret
ss          = Decapsulate(sk, ct) # Decrypt to recover secret
\end{lstlisting}

Compared to classical schemes like RSA, Kyber offers more compact key and ciphertext sizes at higher security levels. Table~\ref{tab:kyber_sizes} compares the relevant sizes in bytes for Kyber and RSA \cite{kyber2024}.

\begin{table}[h]
\centering
\caption{Key and Ciphertext Sizes (bytes)}
\label{tab:kyber_sizes}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|l|c|c|c|}
\hline
\textbf{Scheme} & \textbf{Private Key} & \textbf{Public Key} & \textbf{Ciphertext} \\
\hline
Kyber-512 & 1632 & 800 & 768 \\
Kyber-768 & 2400 & 1184 & 1088 \\
Kyber-1024 & 3168 & 1568 & 1568 \\
RSA-3072 & 384 & 384 & 384 \\
RSA-15360 & 1920 & 1920 & 1920 \\
\hline
\end{tabular}
\end{table}

The parameters that define Kyber’s security and performance are listed in Table~\ref{tab:kyber-params}. These include the polynomial degree, modulus, and bounds on the error distributions used to introduce noise in the ciphertexts \cite{kyber2021}.

\begin{table}[h]
\centering
\caption{Kyber Parameters}
\label{tab:kyber-params}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Description} \\
\midrule
$n = 256$ & Polynomial degree \\
$k \in \{2,3,4\}$ & Number of polynomials (varies by variant) \\
$q = 3329$ & Prime modulus for coefficients \\
$\eta_1, \eta_2$ & Noise bounds in key/ciphertext \\
$d_u, d_v$ & Compression factors \\
\bottomrule
\end{tabular}
\end{table}

Kyber has been optimized for modern platforms. Figure~\ref{fig:kyber_perf} shows benchmark cycle counts for each variant on an Intel Haswell CPU using AVX2 vector instructions \cite{kyber2024}. These results reflect Kyber’s suitability for high-performance and embedded applications alike.

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=12pt,
    width=12cm,
    height=7cm,
    enlarge x limits=0.2,
    ylabel={Cycles},
    symbolic x coords={Kyber-512,Kyber-768,Kyber-1024},
    xtick=data,
    ymin=0,
    legend style={at={(0.5,-0.15)}, anchor=north, legend columns=-1},
    nodes near coords,
    every node near coord/.append style={
        font=\small,
        rotate=90,   % vertical labels
        anchor=west  % position so they sit neatly above bars
    },
    tick label style={font=\small},
    label style={font=\small},
    legend cell align={left},
    grid=major
]

% KeyGen
\addplot[
    pattern=north east lines,
    pattern color=black
] coordinates {
    (Kyber-512,33856) (Kyber-768,52732) (Kyber-1024,73544)
};

% Encaps
\addplot[
    pattern=dots,
    pattern color=black
] coordinates {
    (Kyber-512,45200) (Kyber-768,67624) (Kyber-1024,97324)
};

% Decaps
\addplot[
    pattern=crosshatch,
    pattern color=black
] coordinates {
    (Kyber-512,34572) (Kyber-768,53156) (Kyber-1024,79128)
};

\legend{KeyGen, Encaps, Decaps}
\end{axis}
\end{tikzpicture}
\caption{Kyber Performance on Haswell CPU (AVX2 Optimized)}
\label{fig:kyber_perf}
\end{figure}

For platforms that lack efficient SHAKE implementations, Kyber also offers a variant called \texttt{Kyber-90s}, which uses AES-256 and SHA2 instead. This allows faster execution on devices with AES hardware acceleration, such as many embedded and mobile processors \cite{kyber2024}.
